---
title: "Problem set 2 — Answer Key"
date: 2025-09-18
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

For these exercises, do not load any packages other than **dslabs**.

```{r}
library(dslabs)
```

## 1

What is the sum of the first 150 positive integers? Use the functions `seq` and `sum` to compute the sum with R for any `n`.

```{r}
n <- 150
sum(seq(1, n))
```

## 2

Load the `murders` dataset from **dslabs**. Use `str` to examine the structure of the `murders` object.

* What are the column names used by the data frame for these five variables: state name, abbreviation, region, population, total murders?
* Show the subset of `murders` showing states with less than 1.2 per 100,000 deaths. Show all variables.

```{r}
str(murders)
```

```{r}
# Column names for the requested variables
c("state", "abb", "region", "population", "total")
```

```{r}
rate <- murders$total / murders$population * 100000
murders[rate < 1.2, ]
```

## 3

Show the subset of `murders` showing states with less than 1.2 per 100,000 deaths and in the Northeast of the US. Don't show the `region` variable.

```{r}
rate <- murders$total / murders$population * 100000
out <- murders[rate < 1.2 & murders$region == "Northeast", ]
out[, names(out) != "region"]
```

## 4

Among states with a murder rate less than 1.2 per 100,000, show the smallest population state (state name, population, and rate).

```{r}
rate <- murders$total / murders$population * 100000
low <- rate < 1.2
i <- which(low)[which.min(murders$population[low])]
data.frame(state = murders$state[i], population = murders$population[i], rate = rate[i])
```

## 5

Show the state with a population of more than 8 million with the lowest murder rate (state name, population, and rate).

```{r}
rate <- murders$total / murders$population * 100000
big <- murders$population > 8e6
i <- which(big)[which.min(rate[big])]
data.frame(state = murders$state[i], population = murders$population[i], rate = rate[i])
```

## 6

Compute the murder rate for each region of the US (total murders / total population * 100,000). Return a data frame with one row per region and columns `region` and `rate`.

```{r}
murders_by_region <- tapply(murders$total, murders$region, sum)
pop_by_region <- tapply(murders$population, murders$region, sum)
rate_by_region <- murders_by_region / pop_by_region * 100000
data.frame(region = names(rate_by_region), rate = as.numeric(rate_by_region))
```

## 7

Create a vector of numbers starting at 5, not passing 60, in increments of 3/8. How many numbers are in the list?

```{r}
v <- seq(from = 5, to = 60, by = 3/8)
length(v)
```

## 8

Make the data frame below and add Celsius temperatures.

```{r}
temp_f <- c(72, 95, 41, 86, 78, 33)
city <- c("Chicago", "Lagos", "Oslo", "Rio de Janeiro", 
          "San Juan", "Toronto")
city_temps <- data.frame(name = city, temperature_f = temp_f)
city_temps
```

```{r}
city_temps$temperature_c <- (city_temps$temperature_f - 32) * 5/9
city_temps
```

## 9

Write a function `euler2` that computes (S_n = \sum_{k=1}^n 1/k^2). Test at n=10 and n=100.

```{r}
euler2 <- function(n) sum(1/(seq(1, n)^2))
euler2(10)
euler2(100)
```

## 10

Plot (S_n) versus (n) for (n=1,\dots,2000) with a horizontal dashed line at (\pi^2/6).

```{r}
n_vals <- 1:2000
S <- sapply(n_vals, euler2)
plot(n_vals, S, type = "l", xlab = "n", ylab = expression(S[n]))
abline(h = pi^2/6, lty = 2)
```

## 11

Use `%in%` and `state.abb` to check which of these are real state abbreviations: AL, AK, AZ, AR, AA.

```{r}
test <- c("AL", "AK", "AZ", "AR", "AA")
test %in% state.abb
```

## 12

Report the one entry that is not an actual abbreviation.

```{r}
test <- c("AL", "AK", "AZ", "AR", "AA")
test[which(!(test %in% state.abb))]
```

## 13

In the `murders` dataset, use `%in%` to show all variables for Florida, California, and New York, in that order.

```{r}
targets <- c("Florida", "California", "New York")
sub <- murders[murders$state %in% targets, ]
sub[match(targets, sub$state), ]
```

## 14

Write `vandermonde_helper2(x, n)` that returns ((1, x, x^2, \dots, x^n)). Show results for x=2, n=6.

```{r}
vandermonde_helper2 <- function(x, n) x^(0:n)
vandermonde_helper2(2, 6)
```

## 15

Simulate Bernoulli data and compute run lengths of consecutive 1s (no loop). Plot distribution. Compare empirical proportions for run lengths 1–8 to geometric theory.

```{r}
n <- 20000
p <- 0.35
set.seed(2025-9-18)
x <- sample(c(0,1), n, prob = c(1 - p, p), replace = TRUE)
```

```{r}
r <- rle(x)
ones_lengths <- r$lengths[r$values == 1]

hist(ones_lengths, breaks = 30, xlab = "Run length of 1s",
     main = "Distribution of run lengths of 1s")

tab <- table(ones_lengths)
emp_counts <- as.numeric(tab[as.character(1:8)])
emp_counts[is.na(emp_counts)] <- 0
emp_probs <- emp_counts / length(ones_lengths)

k <- 1:8
theory_probs <- (1 - p) * p^(k - 1)

data.frame(run_length = k, empirical_prob = emp_probs, theory_prob = theory_probs)
```

## 16

In `murders`, compute the national average murder rate. Label states using `ifelse`:

* "High Crime, High Pop" if rate > national avg and pop > 6 million
* "High Crime, Low Pop" if rate > national avg and pop ≤ 6 million
* "Lower Crime" otherwise
  Then show a table of counts.

```{r}
rate <- murders$total / murders$population * 100000
national_rate <- sum(murders$total) / sum(murders$population) * 100000

high_crime <- rate > national_rate
high_pop <- murders$population > 6e6

labels <- ifelse(high_crime & high_pop, "High Crime, High Pop",
                 ifelse(high_crime & !high_pop, "High Crime, Low Pop",
                        "Lower Crime"))
table(labels)
```

## 17

What is the murder rate of the state that ranks 12th in murder rate (highest to lowest)? Show work using `order` (optionally check with `sort` and `rank`).

```{r}
rate <- murders$total / murders$population * 100000
ord <- order(rate, decreasing = TRUE)
i <- ord[12]
data.frame(rank = 12, state = murders$state[i], rate = rate[i])
```

## 18

Write `compute_harmonic_mean(x)` returning the harmonic mean, but `NA` if any values are zero/negative. Test on `c(1,2,4,8)`.

```{r}
compute_harmonic_mean <- function(x) {
  if (any(x <= 0)) return(NA_real_)
  length(x) / sum(1/x)
}
compute_harmonic_mean(c(1, 2, 4, 8))
```

## 19

Create `safe_divide(x, y)` returning `x/y`, but "Cannot divide by zero" when `y` is 0. Vectorized. Test on `x <- c(10,20,30)` and `y <- c(2,0,5)`.

```{r}
safe_divide <- function(x, y) {
  out <- x / y
  out <- as.character(out)
  out[y == 0] <- "Cannot divide by zero"
  out
}

x <- c(10, 20, 30)
y <- c(2, 0, 5)
safe_divide(x, y)
```

## 20

Write `classify_state_safety(state_name)` based on murder rate:

* "Very Safe" if rate < 1
* "Safe" if 1 ≤ rate < 3
* "Moderate" if 3 ≤ rate < 5
* "High Risk" if rate ≥ 5
  If state not found, return "State not found".

Test on "Vermont", "Texas", "California", "NotAState". Then classify all states with `sapply` and tabulate categories.

```{r}
rate <- murders$total / murders$population * 100000
rate_named <- setNames(rate, murders$state)

classify_state_safety <- function(state_name) {
  if (!(state_name %in% names(rate_named))) return("State not found")
  r <- rate_named[state_name]
  if (r < 1) "Very Safe"
  else if (r < 3) "Safe"
  else if (r < 5) "Moderate"
  else "High Risk"
}

classify_state_safety("Vermont")
classify_state_safety("Texas")
classify_state_safety("California")
classify_state_safety("NotAState")
```

```{r}
cats <- sapply(murders$state, classify_state_safety)
table(cats)
```


