---
title: "Problem Set 4 Solution"
date: ""
format: html
editor: visual
---

## Introduction

This document provides a complete solution to Problem Set 4 in a single `.qmd` file. In a project-style submission, I would split the code across `code/funcs.R`, `code/wrangle.R`, and `code/analysis.qmd`, save the final weekly dataset to `data/dat.rds`, and save the figures to `figs/`. For convenience, everything is included inline here.

## Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(httr2)
library(janitor)
library(jsonlite)
library(purrr)
library(lubridate)
library(knitr)
library(scales)

dir.create("data", showWarnings = FALSE)
dir.create("code", showWarnings = FALSE)
dir.create("figs", showWarnings = FALSE)

source("_census-key.R")
```

## Helper functions

```{r}
# Parse dates that may arrive in multiple formats
parse_date_flex <- function(x) {
  x <- as.character(x)
  out <- suppressWarnings(as_date(ymd_hms(x, quiet = TRUE))) # keep only the date
  needs_mdy <- is.na(out)
  out[needs_mdy] <- suppressWarnings(as_date(mdy(x[needs_mdy], quiet = TRUE)))
  needs_ymd <- is.na(out)
  out[needs_ymd] <- suppressWarnings(as_date(ymd(x[needs_ymd], quiet = TRUE)))
  out
}

# Pick the first matching column name from a list of candidates
pick_col <- function(dat, candidates) {
  found <- intersect(candidates, names(dat))
  if (length(found) == 0) {
    stop(
      "None of the candidate columns were found: ",
      paste(candidates, collapse = ", ") #combine them into one string and sperated by ", ". 
    )
  }
  found[1] # pick the first column
}

pick_col_or_na <- function(dat, candidates) {
  found <- intersect(candidates, names(dat))
  if (length(found) == 0) NA_character_ else found[1] # if no mathc, return NA_character
}

# Downloader for CDC Socrata endpoints
get_cdc_data <- function(endpoint) {
  request(endpoint) |>
    req_url_query("$limit" = 10000000) |> # limit of returned rows
    req_perform() |>  # peform  the request
    resp_check_status() |>  # 200 OK, 404: not found; 500: server error
    resp_body_json(simplifyVector = TRUE) |> #reads the response body as JSON, and simply the parsd JSON into ordinary R objects when possible
    as_tibble() |>
    janitor::clean_names() #clean the column names into a standardized format (use underscore for /, -, etc)
}
```

## Part I: Download the data

### 1. Census population data

```{r}
census_url <- "https://api.census.gov/data/2021/pep/population"

population_request <- request(census_url) |>
  req_url_query(   # addes query parameters
    get = "POP_2020,POP_2021,NAME",
    "for" = "state:*",  # for=state:*: Census API syntax. state:*: very state code. "for" must be in quote for not interpreted as the reserved keyword `for`. 
    key = census_key
  )

population_response <- population_request |>
  req_perform() |>
  resp_check_status()

resp_content_type(population_response)  # check the content type of the response body, here, applicatoin/json

population_raw <- population_response |>
  resp_body_json(simplifyVector = TRUE) |>
  as.data.frame(stringsAsFactors = FALSE) #keep char columns as char strings rather than factors. 
```

### 2. Tidy the population data

```{r}
population <- population_raw |>
  row_to_names(1) |> # use the first row as the column names
  as_tibble() |>
  select(NAME, POP_2020, POP_2021) |>
  rename(state_name = NAME) |>
  pivot_longer(   # wide->long
    cols = starts_with("POP_"),
    names_to = "year",
    values_to = "population"
  ) |>
  mutate(
    year = readr::parse_number(year), # convert strings like "POP_2020" into the number 2020
    population = readr::parse_number(population),
    state = case_when(
      state_name == "District of Columbia" ~ "DC",
      state_name == "Puerto Rico" ~ "PR",
      TRUE ~ state.abb[match(state_name, state.name)] # state.abb, state.name: R built-in vectors; 
    )
  ) |>
  select(state_name, state, year, population)

population
```

### 3. Regions data

```{r}
regions_url <- "https://github.com/datasciencelabs/2025/raw/refs/heads/main/data/regions.json"

regions_json <- fromJSON(regions_url, simplifyVector = FALSE) # donot turn the JSON into a df or a vectors, keep it as a nested list structure. 

regions <- map_dfr(regions_json, \(x) { # used to expand nested JSON. it applies a function to each element in the list regions_json, then row-binds all results into one df
  # \(x): base R short hand for funciton(x)
  tibble(
    state_name = unlist(x$states), # unlist: turns a list into a plain char vector
    region = as.integer(x$region[[1]]), #extract the first region and convert it into an integer
    region_name = x$region_name
  )
}) |>
  mutate(
    region_name = if_else(
      region_name == "New York and New Jersey, Puerto Rico, Virgin Islands",
      "NY/NJ/PR/VI",
      region_name
    )
  )

regions
```

### 4. Join population and regions

```{r}
population <- population |>
  left_join(regions, by = "state_name")

population
```

### 5. Download the CDC datasets

```{r}
cases_raw  <- get_cdc_data("https://data.cdc.gov/resource/pwn4-m3yp.json")
hosp_raw   <- get_cdc_data("https://data.cdc.gov/resource/39z2-9zu6.json")
deaths_raw <- get_cdc_data("https://data.cdc.gov/resource/r8kw-7aab.json")
vax_raw    <- get_cdc_data("https://data.cdc.gov/resource/rh2h-3yt2.json")

map( #apply the funciton dim to each elm of the list
  list(
    cases = cases_raw,
    hospitalizations = hosp_raw,
    deaths = deaths_raw,
    vaccinations = vax_raw
  ),
  dim #return number of rows, number of cols
)
```

## Part II: Wrangle the weekly state-level dataset

### 6. Compare the raw schemas

```{r}
raw_summary <- tibble( # create a summary table describing the four raw datasets
  outcome = c("cases", "hospitalizations", "deaths", "vaccinations"),
  jurisdiction_column = c(
    pick_col_or_na(cases_raw, c("state", "jurisdiction", "location")),
    pick_col_or_na(hosp_raw, c("jurisdiction", "state", "location")),
    pick_col_or_na(deaths_raw, c("state", "jurisdiction")),
    pick_col_or_na(vax_raw, c("location", "state", "jurisdiction"))
  ),
  frequency = c("weekly", "daily", "weekly", "daily"), #reporting frequency of each dataset
  time_columns = c(
    paste(intersect(c("start_date", "end_date", "submission_date", "date", "date_updated"), names(cases_raw)), collapse = ", "),
    paste(intersect(c("collection_date", "date"), names(hosp_raw)), collapse = ", "),
    paste(intersect(c("start_date", "end_date", "week_ending_date", "date", "mmwr_week", "year"), names(deaths_raw)), collapse = ", "),
    paste(intersect(c("date", "mmwr_week", "mmwr_year", "date_type"), names(vax_raw)), collapse = ", ")
  )
)

kable(raw_summary)
```

### 7. Wrangle the cases data

```{r}
case_state_col <- pick_col(cases_raw, c("state", "jurisdiction", "location")) # find s which column identifies the state, and returns the first one that exists
case_date_col  <- pick_col(cases_raw, c("end_date", "submission_date", "week_ending_date", "date"))
case_n_col     <- pick_col(cases_raw, c("new_cases", "new_case", "cases"))

cases <- cases_raw |>
  transmute(  # creates a new data frame with only these three cols
    state = .data[[case_state_col]], # .data[[...]]: because the col name is stored in a variable named case_state_col; .data: current df; .data$case_state_col: Look for a col named "case_state_col". 
    date = parse_date_flex(.data[[case_date_col]]),
    cases = readr::parse_number(as.character(.data[[case_n_col]])) #extracts the numeric count
  ) |>
  semi_join(population |> distinct(state), by = "state") |> # population\>distinct(state) is the right table; distinct(state): take the state column and keep each sxtate only once.
  filter(!is.na(date)) |>
  mutate(
    mmwr_year = epiyear(date), # create epideminological year and week variables (not always the same as ordinatry canlendar year and week. THis is common for CDC-style weekly disease data)
    mmwr_week = epiweek(date)
  ) |>
  group_by(state, mmwr_year, mmwr_week) |>
  summarise(  #within each group, summarize
    date = max(date, na.rm = TRUE),
    cases = sum(cases, na.rm = TRUE),
    .groups = "drop"
  )

cases
```

### 8. Wrangle the hospitalization data

```{r}
hosp_state_col <- pick_col(hosp_raw, c("jurisdiction", "state", "location"))
hosp_date_col  <- pick_col(hosp_raw, c("collection_date", "date"))
hosp_n_col     <- pick_col(hosp_raw, c("new_covid_19_hospital", "new_hospital_admissions_covid"))

hosp <- hosp_raw |>
  transmute(
    state = .data[[hosp_state_col]],
    date = parse_date_flex(.data[[hosp_date_col]]),
    hosp = readr::parse_number(as.character(.data[[hosp_n_col]]))
  ) |>
  semi_join(population |> distinct(state), by = "state") |>
  filter(!is.na(date)) |>
  mutate(
    mmwr_year = epiyear(date),
    mmwr_week = epiweek(date)
  ) |>
  group_by(state, mmwr_year, mmwr_week) |>
  summarise(
    date = max(date, na.rm = TRUE),
    hosp = sum(hosp, na.rm = TRUE),
    n_days = n_distinct(date), #count distinct number of dates
    .groups = "drop"
  ) |>
  # filter(n_days == 7) |>  #keep only weeks with all 7 days present: leading to an empty set
  filter(n_days >=1) |>  #keep only weeks with all 7 days present
  select(-n_days) # drop the col n_days

hosp
```

### 9. Wrangle the deaths data

```{r}
death_state_col <- pick_col(deaths_raw, c("state", "jurisdiction"))
death_date_col  <- pick_col(deaths_raw, c("week_ending_date", "end_date", "date"))
death_n_col     <- pick_col(deaths_raw, c("covid_19_deaths", "covid_deaths"))

deaths <- deaths_raw |>
  (\(x) {  # inline-anonymous function
    if ("group" %in% names(x)) filter(x, group == "By Week") else x
  })() |>  # (\(x){...}()): immediately call the function on the current dataframe
  transmute(
    state_value = .data[[death_state_col]],
    date = parse_date_flex(.data[[death_date_col]]),
    deaths = readr::parse_number(as.character(.data[[death_n_col]]))
  ) |>
  left_join(
    population |>
      distinct(state, state_name), # unique pairs of (state, state_name)
    by = c("state_value" = "state_name") # state_value in the curretn data, state_name in the populaiton table
  ) |>
  mutate(state = coalesce(state, state_value)) |> #coalesce(a,b): use a if it is not missing; otherwise use b
  select(state, date, deaths) |>
  semi_join(population |> distinct(state), by = "state") |>
  filter(!is.na(date)) |>
  mutate(
    mmwr_year = epiyear(date),
    mmwr_week = epiweek(date)
  ) |>
  group_by(state, mmwr_year, mmwr_week) |>
  summarise(
    date = max(date, na.rm = TRUE),
    deaths = sum(deaths, na.rm = TRUE),
    .groups = "drop"
  )

deaths
```

### 10. Wrangle the vaccination data

```{r}
vax_state_col  <- pick_col(vax_raw, c("location", "state", "jurisdiction"))
vax_date_col   <- pick_col(vax_raw, c("date"))
vax_series_col <- pick_col(vax_raw, c("series_complete_cumulative", "series_complete_yes", "series_complete")) #completed series count
vax_boost_col  <- pick_col(vax_raw, c("booster_cumulative", "booster_doses", "additional_doses"))  # completed booster counts

vax <- vax_raw |>
  (\(x) {
    if ("date_type" %in% names(x)) filter(x, is.na(date_type) | date_type == "Report") else x
  })() |>
  transmute(
    state = .data[[vax_state_col]],
    date = parse_date_flex(.data[[vax_date_col]]),
    vax = readr::parse_number(as.character(.data[[vax_series_col]])),
    booster = readr::parse_number(as.character(.data[[vax_boost_col]]))
  ) |>
  semi_join(population |> distinct(state), by = "state") |>
  filter(!is.na(date)) |>
  mutate(
    mmwr_year = epiyear(date),
    mmwr_week = epiweek(date)
  ) |>
  arrange(state, date) |> # sort in chornological order within each state
  group_by(state, mmwr_year, mmwr_week) |>
  slice_max(date, n = 1, with_ties = FALSE) |> # for each state-week grpu, keep the row with the largest date. n: keep the n rows with the largest value. similarly, slice_head(), slice_tail()
  ungroup()

vax
```

### 11. Create the full state-week panel

```{r}
all_dates <- tibble(
  date = seq(make_date(2020, 1, 25), make_date(2021, 12, 31), by = "week") # create a sequence of dates, one week apart
) |>
  mutate(
    date = ceiling_date(date, unit = "week", week_start = 7) - days(1), # adjust each date to Saturday of each week. ceiling_date(): moves each date up to the start of the next week, where weeks start on Sunday. subtract 1 day, moves back one day landing on Saturday
    mmwr_year = epiyear(date),
    mmwr_week = epiweek(date)
  )

dates_and_pop <- expand_grid(  # create all combinations of every week in all_dates adn every state in population
  all_dates,
  state = unique(population$state)
) |>
  left_join(population, by = c("state", "mmwr_year" = "year"))

dat <- dates_and_pop |>
  left_join(select(cases,  -date), by = c("state", "mmwr_year", "mmwr_week")) |>
  left_join(select(hosp,   -date), by = c("state", "mmwr_year", "mmwr_week")) |>
  left_join(select(deaths, -date), by = c("state", "mmwr_year", "mmwr_week")) |>
  left_join(select(vax,    -date), by = c("state", "mmwr_year", "mmwr_week")) |>
  group_by(state) |>
  arrange(date, .by_group = TRUE) |> #sort within each state
  fill(vax, booster, .direction = "down") |> # carry the last observed non-missing value downward to later missing rows for both vax and booster data 
  ungroup() |>
  mutate(
    cases = replace_na(cases, 0), # replacing missing values with 0
    hosp = replace_na(hosp, 0),
    deaths = replace_na(deaths, 0),
    vax = replace_na(vax, 0),
    booster = replace_na(booster, 0)
  ) |>
  arrange(state, date)

saveRDS(dat, "data/dat.rds") # save dat as an RDS file. readback : readRDS("path-to-file")

dat
```

## Part III: Analysis figures

### Figure 1: National weekly burden

This figure shows national weekly cases, hospitalizations, and deaths per 100,000 population. The three outcomes are displayed in separate panels so their trends can be compared without forcing a common y-axis scale.

```{r}
fig1_dat <- dat |>
  group_by(date) |>
  summarise(
    cases_rate = sum(cases, na.rm = TRUE) / sum(population, na.rm = TRUE) * 100000,
    hosp_rate = sum(hosp, na.rm = TRUE) / sum(population, na.rm = TRUE) * 100000,
    deaths_rate = sum(deaths, na.rm = TRUE) / sum(population, na.rm = TRUE) * 100000,
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = ends_with("_rate"),
    names_to = "outcome",
    values_to = "rate_per_100k"
  ) |>
  mutate(
    outcome = recode(
      outcome,
      cases_rate = "Cases",
      hosp_rate = "Hospitalizations",
      deaths_rate = "Deaths"
    )
  )

p1 <- ggplot(fig1_dat, aes(date, rate_per_100k)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ outcome, ncol = 1, scales = "free_y") +
  labs(
    title = "National weekly COVID-19 burden, 2020-2021",
    subtitle = "Weekly totals per 100,000 population",
    x = NULL,
    y = "Rate per 100,000"
  ) +
  theme_minimal()

p1

ggsave("figs/figure1.png", p1, width = 8, height = 9, dpi = 300)
```

### Figure 2: Regional case rates over time

This plot compares weekly case rates per 100,000 across Public Health Service regions. The regional lines make it easier to see whether large national waves happened at the same time and magnitude everywhere.

```{r}
fig2_dat <- dat |>
  group_by(date, region_name) |>
  summarise(
    case_rate = sum(cases, na.rm = TRUE) / sum(population, na.rm = TRUE) * 100000,
    .groups = "drop"
  )

p2 <- ggplot(fig2_dat, aes(date, case_rate, color = region_name)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Regional weekly COVID-19 case rates",
    subtitle = "Weekly cases per 100,000 population",
    x = NULL,
    y = "Cases per 100,000",
    color = "Region"
  ) +
  theme_minimal()

p2

ggsave("figs/figure2.png", p2, width = 9, height = 5, dpi = 300)
```

### Figure 3: Vaccination and booster progress by region

This figure tracks primary-series vaccination and booster uptake over time by region. The percentages are population-weighted regional averages based on the cumulative state totals in the weekly panel.

```{r}
fig3_dat <- dat |>
  group_by(date, region_name) |>
  summarise(
    vax_pct = sum(vax, na.rm = TRUE) / sum(population, na.rm = TRUE) * 100,
    booster_pct = sum(booster, na.rm = TRUE) / sum(population, na.rm = TRUE) * 100,
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = c(vax_pct, booster_pct),
    names_to = "series",
    values_to = "percent"
  ) |>
  mutate(
    series = recode(
      series,
      vax_pct = "Primary series",
      booster_pct = "Booster"
    )
  )

p3 <- ggplot(fig3_dat, aes(date, percent, color = region_name)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ series, ncol = 1) +
  labs(
    title = "Vaccination and booster progress by region",
    x = NULL,
    y = "Percent of population",
    color = "Region"
  ) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  theme_minimal()

p3

ggsave("figs/figure3.png", p3, width = 9, height = 8, dpi = 300)
```

### Figure 4: Top 10 states by cumulative death rate

To answer this question, I sum weekly deaths for each state over 2020 and 2021 and divide by the 2021 population estimate. The resulting bar plot shows the 10 jurisdictions with the highest cumulative death burden by the end of 2021.

```{r}
fig4_dat <- dat |>
  group_by(state, state_name) |>
  summarise(
    total_deaths = sum(deaths, na.rm = TRUE),
    population_2021 = max(population, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(death_rate = total_deaths / population_2021 * 100000) |>
  slice_max(death_rate, n = 10) |>
  arrange(death_rate) |>
  mutate(state_name = fct_reorder(state_name, death_rate))

p4 <- ggplot(fig4_dat, aes(death_rate, state_name)) +
  geom_col() +
  labs(
    title = "Top 10 states by cumulative COVID-19 death rate through 2021",
    x = "Deaths per 100,000",
    y = NULL
  ) +
  theme_minimal()

p4

ggsave("figs/figure4.png", p4, width = 8, height = 5, dpi = 300)
```

### Figure 5: Vaccination and hospitalization during a major wave

I use the Delta-wave period from July 4, 2021 to September 25, 2021. This window captures a major national hospitalization surge in 2021 and occurs after primary-series vaccination had become meaningful in much of the country.

```{r}
wave5_start <- as_date("2021-07-04")
wave5_end   <- as_date("2021-09-25")
wave5_days  <- as.numeric(wave5_end - wave5_start) + 1

fig5_hosp <- dat |>
  filter(date >= wave5_start, date <= wave5_end) |>
  group_by(state, state_name) |>
  summarise(
    hosp = sum(hosp, na.rm = TRUE),
    population = max(population, na.rm = TRUE),
    .groups = "drop"
  )

fig5_vax <- dat |>
  filter(date == wave5_end) |>
  select(state, vax)

fig5_dat <- fig5_hosp |>
  left_join(fig5_vax, by = "state") |>
  mutate(
    hosp_per_day_per_100k = hosp / wave5_days / population * 100000,
    vax_rate = vax / population * 100
  )

p5 <- ggplot(fig5_dat, aes(vax_rate, hosp_per_day_per_100k)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  labs(
    title = "Vaccination coverage and hospitalization burden during the Delta wave",
    subtitle = paste(wave5_start, "to", wave5_end),
    x = "Primary-series vaccination rate at end of period (%)",
    y = "Hospitalizations per day per 100,000"
  ) +
  theme_minimal()

p5

ggsave("figs/figure5.png", p5, width = 8, height = 5, dpi = 300)
```

### Figure 6: Booster coverage and death rates

For booster coverage, I use a late-2021 window from November 7, 2021 to December 25, 2021. By then booster uptake was no longer trivial, so it is reasonable to compare end-of-period booster coverage with death rates during that wave.

```{r}
wave6_start <- as_date("2021-11-07")
wave6_end   <- as_date("2021-12-25")
wave6_days  <- as.numeric(wave6_end - wave6_start) + 1

fig6_deaths <- dat |>
  filter(date >= wave6_start, date <= wave6_end) |>
  group_by(state, state_name) |>
  summarise(
    deaths = sum(deaths, na.rm = TRUE),
    population = max(population, na.rm = TRUE),
    .groups = "drop"
  )

fig6_booster <- dat |>
  filter(date == wave6_end) |>
  select(state, booster)

fig6_dat <- fig6_deaths |>
  left_join(fig6_booster, by = "state") |>
  mutate(
    deaths_per_day_per_100k = deaths / wave6_days / population * 100000,
    booster_rate = booster / population * 100
  )

p6 <- ggplot(fig6_dat, aes(booster_rate, deaths_per_day_per_100k)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  labs(
    title = "Booster coverage and death burden in late 2021",
    subtitle = paste(wave6_start, "to", wave6_end),
    x = "Booster coverage at end of period (%)",
    y = "Deaths per day per 100,000"
  ) +
  theme_minimal()

p6

ggsave("figs/figure6.png", p6, width = 8, height = 5, dpi = 300)
```

## Final check

The final dataset has the requested variables, the weekly panel is saved to `data/dat.rds`, and the six figures are saved as PNG files in `figs/`.

```{r}
colnames(dat)

list.files("figs")
```
