---
title: Problem set 5 solutions
date: ""
format: html
execute:
  echo: true
  warning: false
  message: false
---

This solution set explores probability through two settings: repeated dice rolls and American roulette. It uses both Monte Carlo simulation and exact mathematical calculations.

```{r setup}
library(tidyverse)

theme_set(theme_minimal())
options(digits = 4)

# Event: in n rolls of two fair dice, do we see at least one double-six?
double_six <- function(n){
  die1 <- sample(1:6, n, replace = TRUE)
  die2 <- sample(1:6, n, replace = TRUE)
  any(die1 == 6 & die2 == 6)
}

# Exact probability of at least one double-six in n rolls.
exact_prob <- function(n){
  1 - (35/36)^n
}

# Faster Monte Carlo for the double-six event.
# Each roll of two dice is one of 36 equally likely outcomes,
# and one of those outcomes is a double-six.
mc_prob_fast <- function(n, B = 1000){
  sims <- matrix(sample.int(36, n * B, replace = TRUE), nrow = B) #B by n
  mean(rowSums(sims == 36) > 0) # proportion of simulations where each has at least one 36. 
}

# Casino profit from one $1 bet on red:
# -1 if the player wins (ball lands on red), +1 otherwise.
simulate_X <- function(n = 1){
  sample(c(-1, 1), size = n, replace = TRUE, prob = c(18/38, 20/38))
}

# Total casino profit S from n independent $1 bets on red.
# If W is the number of casino wins among n bets, then total profit S = W-(n-W) = 2W - n.
simulate_S <- function(n, B = 10000){
  2 * rbinom(B, size = n, prob = 20/38) - n # rbinom(B, n, 20/38): generate B values of W
}

```
For example
```{r}
set.seed(1)
S <- simulate_S(1000, B = 10000)
mean(S < 0)     # estimated probability casino loses money
mean(S)         # estimated expected profit
sd(S)           # estimated standard error of S
```

## Part I: Rolling two dice

### 1. Function `double_six`

We simulate `n` rolls of two fair dice and return `TRUE` if at least one roll is `(6, 6)`.

```{r q1}
double_six
```

### 2. Estimated probability for 24 rolls

We estimate the probability of seeing at least one double-six in 24 rolls using `B = 1000` Monte Carlo trials.

```{r q2}
set.seed(2026)
B <- 10^3
n <- 24
prob_est_24 <- mean(replicate(B, double_six(n)))
prob_est_24
```

The Monte Carlo estimate is `r round(prob_est_24, 4)`.

### 3. When does the probability exceed 50%?

We repeat the simulation for several values of `n`.

```{r q3}
set.seed(1997)
compute_prob <- function(n, B = 10^3){
  mean(replicate(B, double_six(n)))
}

n_vals <- 1:60
mc_probs <- sapply(n_vals, compute_prob) # sapply(X, FUN) applies FUN to each element of X; lapply returns a list
n_50 <- min(n_vals[mc_probs > 0.5])
n_50
```

From this simulation, the probability first exceeds 50% at about `r n_50` rolls.

For comparison, the exact threshold solves

$$
1 - \left(\frac{35}{36}\right)^n > 0.5.
$$

This is equivalent to

$$
\left(\frac{35}{36}\right)^n < 0.5,
$$

so

$$
n > \frac{\log(0.5)}{\log(35/36)} \approx 24.61.
$$

Therefore the smallest integer is

$$
n = 25.
$$

### 4. Compare Monte Carlo to the exact probability

The exact probability is

$$
\Pr(\text{at least one double-six in } n \text{ rolls})
= 1 - \left(\frac{35}{36}\right)^n.
$$

```{r q4}
plot_df <- tibble(
  n = n_vals,
  monte_carlo = mc_probs,
  exact = exact_prob(n_vals)
)

ggplot(plot_df, aes(n, monte_carlo)) + # ggplot sets the global aes(x=, y=)
  geom_point(size = 1.8) +
  geom_line(aes(y = exact), color = "red", linewidth = 1) + #overide y locally
  labs(
    title = "Double-six probability: Monte Carlo vs exact",
    x = "Number of rolls (n)",
    y = "Probability"
  )
```

The red curve gives the exact probability, while the points are the Monte Carlo estimates based on `B = 1000` trials.

### 5. How large should `B` be?

Now we fix `n = 24` and vary the number of Monte Carlo iterations.

```{r q5}
set.seed(1998)
B_seq <- seq(10, 250, 5)^2
n <- 24
exact_p24 <- exact_prob(n)
est_probs_q5 <- sapply(B_seq, function(b) mc_prob_fast(n, b))

within_band <- abs(est_probs_q5 - exact_p24) <= 0.005
threshold_candidates <- which(vapply(seq_along(est_probs_q5), function(i) {
  all(within_band[i:length(within_band)]) # check for each index i to the end, whether all remaining values are true
}, logical(1))) #vapply is safer than sapply, it rquires to speicfy the output type. logical(1): each funcition must return exactly one logical value. 
threshold_index <- if (length(threshold_candidates) == 0) length(B_seq) else threshold_candidates[1]
# if no such index, use the lat one, otherwise use the first valid threshold

sqrt_B_threshold <- sqrt(B_seq[threshold_index]) # find the first value of  sqrt(B) after which the MC estimates stays within the desired error band

q5_df <- tibble(
  sqrt_B = sqrt(B_seq),
  estimate = est_probs_q5
)

ggplot(q5_df, aes(sqrt_B, estimate)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = exact_p24, color = "red", linewidth = 1) +
  geom_hline(yintercept = exact_p24 + 0.005, linetype = 2, color = "blue") +
  geom_hline(yintercept = exact_p24 - 0.005, linetype = 2, color = "blue") +
  labs(
    title = "Monte Carlo accuracy for n = 24",
    x = expression(sqrt(B)),
    y = "Estimated probability"
  )
```

In this run, the estimates stay within `0.005` of the exact probability from about $\sqrt(B)$ = `r sqrt_B_threshold` onward. This corresponds to B = `r B_seq[threshold_index]`.

### 6. Improved Monte Carlo versus exact curve

Based on question 5, a value around 40,000 iterations is enough to make the Monte Carlo points practically fall on the exact curve while still rendering quickly.

```{r q6}
set.seed(1999)
B_star <- max(40000, B_seq[threshold_index])
n_vals <- 1:60
mc_probs_better <- sapply(n_vals, function(n) mc_prob_fast(n, B_star))

plot_df_better <- tibble(
  n = n_vals,
  monte_carlo = mc_probs_better,
  exact = exact_prob(n_vals)
)

ggplot(plot_df_better, aes(n, monte_carlo)) +
  geom_point(size = 1.5) +
  geom_line(aes(y = exact), color = "red", linewidth = 1) +
  labs(
    title = paste("Monte Carlo vs exact with B =", format(B_star, big.mark = ",")),
    x = "Number of rolls (n)",
    y = "Probability"
  )
```

## Part II: American roulette

In American roulette there are 38 slots total: 18 red, 18 black, and 2 green.

### 7. Probability that the casino wins a $1 bet on red

If a player bets on red, then the casino wins whenever the outcome is black or green.

There are

- 18 black slots,
- 2 green slots,

so the casino wins on

$$
18 + 2 = 20
$$

of the 38 slots. Therefore,

$$
\Pr(\text{casino wins}) = \frac{20}{38} = \frac{10}{19}.
$$

### 8. Sampling model for $X$

Here $X$ is the casino's profit from a single $1 bet on red:

- $X = -1$ if the ball lands on red,
- $X = +1$ otherwise.

```{r q8}
set.seed(2030)
X <- simulate_X(10)
X
```

So one convenient sampling model is

```{r q8-model}
sample(c(-1, 1), size = 1, replace = TRUE, prob = c(18/38, 20/38))
```

### 9. Probability that the casino loses money when 1000 people play

Let $S$ be the casino's total profit from `n = 1000` independent bets on red.

```{r q9}
set.seed(2031)
n <- 1000
B <- 10^4
S <- simulate_S(n, B)
prob_lose_mc <- mean(S < 0)
prob_lose_mc
```

The Monte Carlo estimate of the probability that the casino loses money is `r round(prob_lose_mc, 4)`.

### 10. Expected value of $X$

Since

$$
\Pr(X = -1) = \frac{18}{38}
\qquad \text{and} \qquad
\Pr(X = 1) = \frac{20}{38},
$$

we have

$$
E[X] = (-1)\left(\frac{18}{38}\right) + (1)\left(\frac{20}{38}\right)
= \frac{2}{38}
= \frac{1}{19}.
$$

Therefore,

$$
E[X] = \frac{1}{19} \approx 0.0526.
$$

### 11. Standard error of $X$

Here the "standard error" of the random variable $X$ is its standard deviation.

First,

$$
E[X^2] = (-1)^2\left(\frac{18}{38}\right) + (1)^2\left(\frac{20}{38}\right) = 1.
$$

So

$$
\operatorname{Var}(X) = E[X^2] - (E[X])^2 = 1 - \left(\frac{1}{19}\right)^2
= \frac{360}{361}.
$$

Hence,

$$
\operatorname{SE}(X) = \sqrt{\operatorname{Var}(X)}
= \sqrt{\frac{360}{361}}
= \frac{6\sqrt{10}}{19}
\approx 0.9986.
$$

### 12. Expected value of $S$

Since $S$ is the sum of 1000 independent copies of $X$,

$$
E[S] = nE[X] = 1000\left(\frac{1}{19}\right) = \frac{1000}{19} \approx 52.63.
$$

```{r q12}
expected_S <- n / 19
mc_mean_S <- mean(S)
c(expected_S = expected_S, monte_carlo_mean = mc_mean_S)
```

The Monte Carlo mean is very close to the theoretical value, so the simulation confirms the result.

### 13. Standard error of $S$

Because $S$ is the sum of `n = 1000` independent copies of $X$,

$$
\operatorname{Var}(S) = n\operatorname{Var}(X)
= 1000\left(\frac{360}{361}\right).
$$

Therefore,

$$
\operatorname{SE}(S) = \sqrt{\operatorname{Var}(S)}
= \sqrt{1000}\cdot \frac{6\sqrt{10}}{19}
= \frac{600}{19}
\approx 31.58.
$$

```{r q13}
se_S <- 600 / 19
mc_sd_S <- sd(S)
c(theoretical_se = se_S, monte_carlo_sd = mc_sd_S)
```

Again, the Monte Carlo standard deviation is close to the theoretical value.

### 14. Visualizing the distribution of $S$

We standardize $S$ and then make a histogram and a QQ-plot.

```{r q14}
S_std <- (S - expected_S) / se_S

tmp_par <- par(mfrow = c(1, 2)) # next plots are arranged in a 1 row by 2 columns layout, and saves the preivous graphics seetings in tmp_par which can be restored by using par(tmp_par)

hist(
  S_std,
  breaks = 30,
  main = "Histogram of standardized S",
  xlab = "Standardized S"
)

qq_df <- data.frame(
  theoretical = qnorm(ppoints(length(S_std))),
  sample = sort(S_std)
)
plot(qq_df$theoretical, qq_df$sample,
     main = "QQ-plot of standardized S",
     xlab = "Theoretical quantiles",
     ylab = "Sample quantiles")
abline(0, 1, col = "red", lwd = 2) # line y=x, intercept=0, slope=1

par(tmp_par) #restore the preivous plotting setting
```

The histogram is close to bell-shaped, and the QQ-plot lies close to the line $y = x$, so the normal approximation is reasonable.

### 15. What improves the normal approximation?

Increasing the number of people playing `n` improves the normal approximation more. The Central Limit Theorem becomes more accurate as the number of independent terms in the sum increases. Increasing the number of Monte Carlo iterations `B` only makes the simulation estimate more stable; it does not make the true distribution of $S$ more normal.

### 16. CLT approximation for the probability the casino loses money

From the previous parts,

$$
E[S] = \frac{n}{19},
\qquad
\operatorname{SE}(S) = \frac{6\sqrt{10n}}{19}.
$$

So by the Central Limit Theorem,

$$
\Pr(S < 0)
\approx
\Phi\left(
\frac{0 - n/19}{6\sqrt{10n}/19}
\right).
$$

For `n = 1000`, this becomes

$$
\Pr(S < 0)
\approx
\Phi\left(
\frac{-1000/19}{6\sqrt{10000}/19}
\right)
=
\Phi\left(-\frac{5}{3}\right).
$$

```{r q16}
clt_prob <- pnorm(0, mean = expected_S, sd = se_S)
clt_prob_cc <- pnorm(-1, mean = expected_S, sd = se_S)  # continuity correction. S=2W-n, taking on even values, P(S<0)=P(S<=-2) \approx P(Y\le -1) where Y is a normal approximation. use -1 because the boundary between the included value -2 and the excluded value 0 is -1. 
c(monte_carlo = prob_lose_mc, clt = clt_prob, clt_continuity_corrected = clt_prob_cc)
```

The plain CLT approximation is close to the Monte Carlo estimate, and the continuity-corrected version is usually a little closer.

### 17. Minimum number of players so the casino loses with probability about 1%

We want

$$
\Pr(S < 0) \approx 0.01.
$$

Using the CLT,

$$
\Phi\left(
\frac{0 - n/19}{6\sqrt{10n}/19}
\right)
\approx 0.01.
$$

So we set

$$
\frac{0 - n/19}{6\sqrt{10n}/19} = z_{0.01}.
$$

Since $z_{0.01} = -z_{0.99}$, this gives

$$
\frac{\sqrt{n}}{6\sqrt{10}} = z_{0.99}.
$$

Hence,

$$
n = \left(6\sqrt{10}\, z_{0.99}\right)^2.
$$

```{r q17}
n_star <- ceiling((6 * sqrt(10) * qnorm(0.99))^2)
n_star
```

Thus the CLT gives n = `r n_star`.

Now we check with Monte Carlo simulation.

```{r q17-check}
set.seed(2032)
B_check <- 10^5
prob_n_star <- mean(simulate_S(n_star, B_check) < 0)
prob_prev <- mean(simulate_S(n_star - 1, B_check) < 0)

c(
  `n_star - 1` = prob_prev,
  n_star = prob_n_star
)
```

The simulation shows that n = `r n_star` brings the casino's chance of losing to about 1%, while n = `r n_star - 1` is still slightly above that level.
