---
title: "Problem Set 5 Solution"
date: "2025-10-12"
format: html
editor: visual
---

## Introduction

This document provides a complete solution in one `.qmd` file. In a project-style submission, I would place the helper function in `code/funcs.R`, the wrangling code in `code/wrangle.R`, save the final dataset to `data/dat.rds`, and save figures to the `figs` directory. For convenience, all of that code is included inline here.

## Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(httr2)
library(janitor)
library(jsonlite)
library(purrr)
library(lubridate)
library(knitr)
library(scales)

dir.create("data", showWarnings = FALSE)
dir.create("code", showWarnings = FALSE)
dir.create("figs", showWarnings = FALSE)

# source("census-key.R")
```

```{r}
# Helper used repeatedly to parse dates that may arrive in different formats
to_date <- function(x) {
  x <- as.character(x)
  out <- suppressWarnings(as_date(ymd_hms(x, quiet = TRUE)))
  needs_mdy <- is.na(out)
  out[needs_mdy] <- suppressWarnings(as_date(mdy(x[needs_mdy], quiet = TRUE)))
  needs_ymd <- is.na(out)
  out[needs_ymd] <- suppressWarnings(as_date(ymd(x[needs_ymd], quiet = TRUE)))
  out
}

# A small helper for choosing the first matching column name
pick_col <- function(dat, candidates) {
  found <- intersect(candidates, names(dat))
  if (length(found) == 0) {
    stop("None of the candidate columns were found: ",
         paste(candidates, collapse = ", "))
  }
  found[1]
}

pick_col_or_na <- function(dat, candidates) {
  found <- intersect(candidates, names(dat))
  if (length(found) == 0) NA_character_ else found[1]
}
```

## 1. Create the population data frame

```{r}
census_key <- "ec6df7cd37913b2b5356e95e373be4fb9befb5d0"
census_url <- "https://api.census.gov/data/2021/pep/population"

population_request <- request(census_url) |>
  req_url_query(
    get = "POP_2020,POP_2021,NAME",
    "for" = "state:*",
    key = census_key
  )

population_response <- population_request |>
  req_perform() |>
  resp_check_status()

population_raw <- population_response |>
  resp_body_json(simplifyVector = TRUE) |>
  as.data.frame(stringsAsFactors = FALSE)

population <- population_raw |>
  row_to_names(1) |>
  as_tibble() |>
  select(NAME, POP_2020, POP_2021) |>
  rename(state_name = NAME) |>
  pivot_longer(
    cols = starts_with("POP_"),
    names_to = "year",
    values_to = "population"
  ) |>
  mutate(
    year = readr::parse_number(year),
    population = readr::parse_number(population),
    state = case_when(
      state_name == "District of Columbia" ~ "DC",
      state_name == "Puerto Rico" ~ "PR",
      TRUE ~ state.abb[match(state_name, state.name)]
    )
  )
```

## 2. Read in regions data and join it to population

```{r}
regions_url <- "https://github.com/datasciencelabs/2025/raw/refs/heads/main/data/regions.json"

regions_json <- fromJSON(regions_url, simplifyVector = FALSE)

regions <- map_dfr(regions_json, \(x) {
  tibble(
    state_name = unlist(x$states),
    region = as.integer(x$region[[1]]),
    region_name = x$region_name
  )
}) |>
  mutate(
    region_name = if_else(
      region_name == "New York and New Jersey, Puerto Rico, Virgin Islands",
      "NY/NJ/PR/VI",
      region_name
    )
  )

population <- population |>
  left_join(regions, by = "state_name")

population
```

## 3. Write a reusable CDC downloader

In a project submission, I would place this function in `code/funcs.R`.

```{r}
get_cdc_data <- function(endpoint) {
  resp <- request(endpoint) |>
    req_url_query("$limit" = 10000000) |>
    req_perform() |>
    resp_check_status()

  resp |>
    resp_body_json(simplifyVector = TRUE) |>
    as_tibble() |>
    clean_names()
}
```

## 4. Download the CDC data

In a project submission, I would place this block in `code/wrangle.R`.

```{r}
cases_raw  <- get_cdc_data("https://data.cdc.gov/resource/pwn4-m3yp.json")
hosp_raw   <- get_cdc_data("https://data.cdc.gov/resource/39z2-9zu6.json")
deaths_raw <- get_cdc_data("https://data.cdc.gov/resource/r8kw-7aab.json")
vax_raw    <- get_cdc_data("https://data.cdc.gov/resource/rh2h-3yt2.json")
```

## 5. Compare the raw variable names across datasets

```{r}
raw_summary <- tibble(
  outcome = c("cases", "hospitalizations", "deaths", "vaccines"),
  jurisdiction_variable_name = c(
    pick_col_or_na(cases_raw,  c("state", "jurisdiction", "location")),
    pick_col_or_na(hosp_raw,   c("jurisdiction", "state", "location")),
    pick_col_or_na(deaths_raw, c("state", "jurisdiction")),
    pick_col_or_na(vax_raw,    c("location", "state", "jurisdiction"))
  ),
  rate = c("weekly", "daily", "weekly", "daily"),
  time_variable_names = c(
    paste(intersect(c("start_date", "end_date", "submission_date", "date", "date_updated"), names(cases_raw)), collapse = ", "),
    paste(intersect(c("collection_date", "date"), names(hosp_raw)), collapse = ", "),
    paste(intersect(c("start_date", "end_date", "week_ending_date", "date"), names(deaths_raw)), collapse = ", "),
    paste(intersect(c("date", "mmwr_week", "mmwr_year"), names(vax_raw)), collapse = ", ")
  )
)

kable(raw_summary)
```

## 6. Wrangle the cases data

```{r}
case_state_col <- pick_col(cases_raw, c("state", "jurisdiction", "location"))
case_date_col  <- pick_col(cases_raw, c("end_date", "submission_date", "week_ending_date", "date"))
case_n_col     <- pick_col(cases_raw, c("new_cases", "new_case", "cases"))

cases <- cases_raw |>
  transmute(
    state = .data[[case_state_col]],
    date = to_date(.data[[case_date_col]]),
    cases = readr::parse_number(as.character(.data[[case_n_col]]))
  ) |>
  semi_join(population |> distinct(state), by = "state") |>
  filter(!is.na(date)) |>
  mutate(
    mmwr_year = epiyear(date),
    mmwr_week = epiweek(date)
  ) |>
  group_by(state, mmwr_year, mmwr_week) |>
  summarise(
    date = max(date, na.rm = TRUE),
    cases = sum(cases, na.rm = TRUE),
    .groups = "drop"
  )

cases
```

## 7. Wrangle the hospitalization data

```{r}
hosp_state_col <- pick_col(hosp_raw, c("jurisdiction", "state", "location"))
hosp_date_col  <- pick_col(hosp_raw, c("collection_date", "date"))
hosp_n_col     <- pick_col(hosp_raw, c("new_covid_19_hospital", "new_hospital_admissions_covid"))

hosp <- hosp_raw |>
  transmute(
    state = .data[[hosp_state_col]],
    date = to_date(.data[[hosp_date_col]]),
    hosp = readr::parse_number(as.character(.data[[hosp_n_col]]))
  ) |>
  semi_join(population |> distinct(state), by = "state") |>
  filter(!is.na(date)) |>
  mutate(
    mmwr_year = epiyear(date),
    mmwr_week = epiweek(date)
  ) |>
  group_by(state, mmwr_year, mmwr_week) |>
  summarise(
    date = max(date, na.rm = TRUE),
    hosp = sum(hosp, na.rm = TRUE),
    n_days = n_distinct(date),
    .groups = "drop"
  ) |>
  filter(n_days == 7) |>
  select(-n_days)

hosp
```

## 8. Wrangle the provisional deaths data

```{r}
death_state_col <- pick_col(deaths_raw, c("state", "jurisdiction"))
death_date_col  <- pick_col(deaths_raw, c("week_ending_date", "end_date", "date"))
death_n_col     <- pick_col(deaths_raw, c("covid_19_deaths", "covid_deaths"))

deaths <- deaths_raw |>
  (\(x) {
    if ("group" %in% names(x)) filter(x, group == "By Week") else x
  })() |>
  transmute(
    state_value = .data[[death_state_col]],
    date = to_date(.data[[death_date_col]]),
    deaths = readr::parse_number(as.character(.data[[death_n_col]]))
  ) |>
  left_join(
    population |>
      distinct(state, state_name),
    by = c("state_value" = "state_name")
  ) |>
  mutate(
    state = coalesce(state, state_value)
  ) |>
  select(state, date, deaths) |>
  semi_join(population |> distinct(state), by = "state") |>
  filter(!is.na(date)) |>
  mutate(
    mmwr_year = epiyear(date),
    mmwr_week = epiweek(date)
  ) |>
  group_by(state, mmwr_year, mmwr_week) |>
  summarise(
    date = max(date, na.rm = TRUE),
    deaths = sum(deaths, na.rm = TRUE),
    .groups = "drop"
  )

deaths
```

## 9. Wrangle the vaccination data

```{r}
vax_state_col   <- pick_col(vax_raw, c("location", "state", "jurisdiction"))
vax_date_col    <- pick_col(vax_raw, c("date"))
vax_series_col  <- pick_col(vax_raw, c("series_complete_cumulative", "series_complete_yes", "series_complete"))
vax_boost_col   <- pick_col(vax_raw, c("booster_cumulative", "booster_doses", "additional_doses"))

vax <- vax_raw |>
  (\(x) {
    if ("date_type" %in% names(x)) filter(x, is.na(date_type) | date_type == "Report") else x
  })() |>
  transmute(
    state = .data[[vax_state_col]],
    date = to_date(.data[[vax_date_col]]),
    vax = readr::parse_number(as.character(.data[[vax_series_col]])),
    booster = readr::parse_number(as.character(.data[[vax_boost_col]]))
  ) |>
  semi_join(population |> distinct(state), by = "state") |>
  filter(!is.na(date)) |>
  mutate(
    mmwr_year = epiyear(date),
    mmwr_week = epiweek(date)
  ) |>
  arrange(state, date) |>
  group_by(state, mmwr_year, mmwr_week) |>
  slice_max(date, n = 1, with_ties = FALSE) |>
  ungroup()

vax
```

## 10. Create the full weekly table and save it

```{r}
all_dates <- data.frame(
  date = seq(make_date(2020, 1, 25), make_date(2021, 12, 31), by = "week")
) |>
  mutate(
    date = ceiling_date(date, unit = "week", week_start = 7) - days(1),
    mmwr_year = epiyear(date),
    mmwr_week = epiweek(date)
  )

dates_and_pop <- cross_join(
  all_dates,
  data.frame(state = unique(population$state))
) |>
  left_join(
    population,
    by = c("state", "mmwr_year" = "year")
  )

dat <- dates_and_pop |>
  left_join(select(cases,  -date), by = c("state", "mmwr_year", "mmwr_week")) |>
  left_join(select(hosp,   -date), by = c("state", "mmwr_year", "mmwr_week")) |>
  left_join(select(deaths, -date), by = c("state", "mmwr_year", "mmwr_week")) |>
  left_join(select(vax,    -date), by = c("state", "mmwr_year", "mmwr_week")) |>
  group_by(state) |>
  arrange(date, .by_group = TRUE) |>
  fill(vax, booster, .direction = "down") |>
  ungroup() |>
  mutate(
    cases = replace_na(cases, 0),
    hosp = replace_na(hosp, 0),
    deaths = replace_na(deaths, 0),
    vax = replace_na(vax, 0),
    booster = replace_na(booster, 0)
  ) |>
  arrange(state, date)

saveRDS(dat, "data/dat.rds")

dat
```

## Figure 1

This figure shows weekly cases, hospitalizations, and deaths, all expressed as rates per 100,000 population.

```{r}
fig1_dat <- dat |>
  group_by(date) |>
  summarise(
    cases_rate = sum(cases, na.rm = TRUE) / sum(population, na.rm = TRUE) * 100000,
    hosp_rate = sum(hosp, na.rm = TRUE) / sum(population, na.rm = TRUE) * 100000,
    deaths_rate = sum(deaths, na.rm = TRUE) / sum(population, na.rm = TRUE) * 100000,
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = ends_with("_rate"),
    names_to = "outcome",
    values_to = "rate_per_100k"
  ) |>
  mutate(
    outcome = recode(
      outcome,
      cases_rate = "Cases",
      hosp_rate = "Hospitalizations",
      deaths_rate = "Deaths"
    )
  )

p1 <- fig1_dat |>
  ggplot(aes(date, rate_per_100k)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ outcome, ncol = 1, scales = "free_y") +
  labs(
    title = "Weekly COVID-19 burden in the United States",
    subtitle = "Rates are weekly totals per 100,000 people",
    x = NULL,
    y = "Rate per 100,000"
  ) +
  theme_minimal()

p1
ggsave("figs/figure1.png", p1, width = 8, height = 9, dpi = 300)
```

## Figure 2

This figure shows the percent of the US population that had completed the primary series and the percent that had received a booster dose.

```{r}
fig2_dat <- dat |>
  group_by(date) |>
  summarise(
    pct_vax = sum(vax, na.rm = TRUE) / sum(population, na.rm = TRUE) * 100,
    pct_booster = sum(booster, na.rm = TRUE) / sum(population, na.rm = TRUE) * 100,
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = c(pct_vax, pct_booster),
    names_to = "series",
    values_to = "percent"
  ) |>
  mutate(
    series = recode(
      series,
      pct_vax = "Primary series",
      pct_booster = "Booster"
    )
  )

p2 <- ggplot(fig2_dat, aes(date, percent, linetype = series)) +
  geom_line(linewidth = 0.8) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(
    title = "Vaccination progress in the United States",
    x = NULL,
    y = "Percent of population",
    linetype = NULL
  ) +
  theme_minimal()

p2
ggsave("figs/figure2.png", p2, width = 8, height = 5, dpi = 300)
```

## Figure 3

This figure describes the distribution of state vaccination rates on July 1, 2021. Because the working dataset is weekly, I use the most recent weekly observation on or before July 1, 2021.

```{r}
ref_date <- max(dat$date[dat$date <= as_date("2021-07-01")])

fig3_dat <- dat |>
  filter(date == ref_date) |>
  mutate(vax_rate = vax / population * 100)

p3 <- ggplot(fig3_dat, aes(vax_rate)) +
  geom_histogram(bins = 15, boundary = 0) +
  labs(
    title = paste("Distribution of vaccination rates by state on", ref_date),
    x = "Primary-series vaccination rate (%)",
    y = "Number of jurisdictions"
  ) +
  theme_minimal()

p3
ggsave("figs/figure3.png", p3, width = 8, height = 5, dpi = 300)
```

## Figure 4

This figure compares regional vaccination progress over time using population-weighted regional averages. The plot typically shows noticeable regional separation, with some regions reaching higher vaccination levels earlier than others.

```{r}
fig4_dat <- dat |>
  group_by(date, region_name) |>
  summarise(
    region_vax_rate = sum(vax, na.rm = TRUE) / sum(population, na.rm = TRUE) * 100,
    .groups = "drop"
  )

p4 <- ggplot(fig4_dat, aes(date, region_vax_rate, color = region_name)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Regional differences in vaccination rates",
    x = NULL,
    y = "Primary-series vaccination rate (%)"
  ) +
  theme_minimal()

p4
ggsave("figs/figure4.png", p4, width = 8, height = 5, dpi = 300)
```

## Figure 5

I use two defensible 2021 windows that combine substantial national COVID-19 activity with meaningful vaccine coverage:

-   July 4, 2021 to September 25, 2021
-   November 7, 2021 to December 25, 2021

For each state and period, I compute deaths per day per 100,000 and the primary-series vaccination rate at the end of the period.

```{r}
periods <- tribble(
  ~period, ~start_date, ~end_date,
  "Summer/Fall 2021 wave", as_date("2021-07-04"), as_date("2021-09-25"),
  "Late 2021 wave",        as_date("2021-11-07"), as_date("2021-12-25")
)

fig5_dat <- periods |>
  mutate(
    data = pmap(
      list(period, start_date, end_date),
      \(period, start_date, end_date) {
        death_part <- dat |>
          filter(date >= start_date, date <= end_date) |>
          group_by(state, population) |>
          summarise(
            deaths = sum(deaths, na.rm = TRUE),
            n_days = as.numeric(max(date) - min(date)) + 7,
            .groups = "drop"
          )

        vax_part <- dat |>
          filter(date == end_date) |>
          select(state, vax, population)

        death_part |>
          left_join(vax_part, by = c("state", "population")) |>
          mutate(
            period = period,
            deaths_per_day_per_100k = deaths / n_days / population * 100000,
            vax_rate = vax / population * 100
          )
      }
    )
  ) |>
  pull(data) |>
  list_rbind()

p5 <- ggplot(fig5_dat, aes(vax_rate, deaths_per_day_per_100k)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ period) +
  labs(
    title = "Vaccination rates and COVID-19 death rates",
    x = "Primary-series vaccination rate (%)",
    y = "Deaths per day per 100,000"
  ) +
  theme_minimal()

p5
ggsave("figs/figure5.png", p5, width = 9, height = 5, dpi = 300)
```

## Figure 6

This repeats the previous exercise using booster coverage instead of primary-series vaccination coverage.

```{r}
fig6_dat <- periods |>
  mutate(
    data = pmap(
      list(period, start_date, end_date),
      \(period, start_date, end_date) {
        death_part <- dat |>
          filter(date >= start_date, date <= end_date) |>
          group_by(state, population) |>
          summarise(
            deaths = sum(deaths, na.rm = TRUE),
            n_days = as.numeric(max(date) - min(date)) + 7,
            .groups = "drop"
          )

        booster_part <- dat |>
          filter(date == end_date) |>
          select(state, booster, population)

        death_part |>
          left_join(booster_part, by = c("state", "population")) |>
          mutate(
            period = period,
            deaths_per_day_per_100k = deaths / n_days / population * 100000,
            booster_rate = booster / population * 100
          )
      }
    )
  ) |>
  pull(data) |>
  list_rbind()

p6 <- ggplot(fig6_dat, aes(booster_rate, deaths_per_day_per_100k)) +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ period) +
  labs(
    title = "Booster coverage and COVID-19 death rates",
    x = "Booster coverage (%)",
    y = "Deaths per day per 100,000"
  ) +
  theme_minimal()

p6
ggsave("figs/figure6.png", p6, width = 9, height = 5, dpi = 300)
```

## Save helper scripts to the requested project structure

The chunks below write out the helper function and wrangling code into the requested `code` directory so the project structure matches the assignment.

```{r}
#| eval: false

# I would save the helper function to code/funcs.R and the wrangling
# pipeline to code/wrangle.R in the final project repository.
```
