---
title: Problem set 4
date: 2025-10-05
format: html
---

```{r setup, include=FALSE}
library(tidyverse)
library(httr2)
library(janitor)
library(jsonlite)
library(purrr)
library(lubridate)
library(knitr)
```

In the next problem set, we plan to explore the relationship between COVID-19 death rates and vaccination rates across US states by visually examining their correlation. This analysis will involve gathering COVID-19 related data from the CDC's API and then extensively processing it to merge the various datasets. Since the population sizes of states vary significantly, we will focus on comparing rates rather than absolute numbers. To facilitate this, we will also source population data from the US Census to accurately calculate these rates.

In this problem set we will learn how to extract and wrangle data from the US Census and CDC APIs.

## 1

Get an API key from the US Census at <https://api.census.gov/data/key_signup.html>. You can't share this public key. But your code has to run on a TFs computer. Assume the TF will have a file in their working directory named `census-key.R` with the following one line of code:

```{r}
# census_key <- "A_CENSUS_KEY_THAT_WORKS"
census_key <- "ec6df7cd37913b2b5356e95e373be4fb9befb5d0"
```

Write a first line of code for your problem set that defines `census_key` by running the code in the file `census-key.R`.

```{r}
# source("census-key.R")
```

## 2

The [US Census API User Guide](https://www.census.gov/content/dam/Census/data/developers/api-user-guide/api-user-guide.pdf) provides details on how to leverage this valuable resource. We are interested in vintage population estimates. From the documentation we find that the _endpoint_ is:

```{r}
url <- "https://api.census.gov/data/2021/pep/population"
```

Use the **httr2** package to construct the following GET request.

```r
https://api.census.gov/data/2021/pep/population?get=POP_2020,POP_2021,NAME&for=state:*&key=YOURKEYHERE
```

Create an object called `request` of class `httr2_request` with this URL as an endpoint.

```{r}
#| message: false
#| warning: false
request <- httr2::request(url) |>
  req_url_query(
    get = "POP_2020,POP_2021,NAME",
    `for` = "state:*",
    key = census_key
  )

request
```

## 3

Make a request to the US Census API using the `request` object. Save the response to an object named `response`. Check the response status of your request and make sure it was successful.

```{r}
response <- request |> req_perform()
resp_status(response)
stopifnot(resp_status(response) == 200)
```

## 4

Use a function from the **httr2** package to determine the content type of your response.

```{r}
resp_content_type(response)
```

## 5

Use just one line of code and one function to extract the data into a matrix.

```{r}
population <- resp_body_json(response, simplifyVector = TRUE)
```

## 6

Examine the `population` matrix you just created. Notice that 1) it is not tidy, 2) the column types are not what we want, and 3) the first row is a header. Convert `population` to a tidy dataset. Remove the state ID column and change the name of the column with state names to `state_name`. Add a column with state abbreviations called `state`. Make sure you assign the abbreviations for DC and PR correctly.

```{r}
population <- population |>
  as.data.frame(stringsAsFactors = FALSE) |>
  row_to_names(row_number = 1) |>
  as_tibble() |>
  select(-state) |>
  rename(state_name = NAME) |>
  pivot_longer(
    cols = starts_with("POP_"),
    names_to = "year",
    values_to = "population"
  ) |>
  mutate(
    year = parse_number(year),
    population = parse_number(population),
    state = case_when(
      state_name == "District of Columbia" ~ "DC",
      state_name == "Puerto Rico" ~ "PR",
      TRUE ~ state.abb[match(state_name, state.name)]
    )
  )

population
```

## 7

As a check, make a barplot of states' populations. Show the state names on the y-axis ordered by population size.

```{r}
population |>
  ggplot(aes(x = reorder(state_name, population), y = population)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ year, scales = "free_y") +
  labs(x = NULL, y = "Population", title = "State populations by year")
```

## 8

The following URL:

```{r}
url <- "https://github.com/datasciencelabs/2025/raw/refs/heads/main/data/regions.json"
```

points to a JSON file that lists the states in the 10 Public Health Service (PHS) regions defined by CDC. We want to add these regions to the `population` dataset. To facilitate this create a data frame called `regions` that has three columns: `state_name`, `region`, and `region_name`. One of the regions has a long name. Change it to something shorter.

```{r}
#| message: false
#| warning: false
regions_raw <- fromJSON(url, simplifyDataFrame = FALSE)

regions <- map_df(
  regions_raw,
  \(x) tibble(
    state_name = x$states,
    region = x$region[[1]],
    region_name = x$region_name
  )
) |>
  mutate(
    region_name = if_else(
      region_name == "New York and New Jersey, Puerto Rico, Virgin Islands",
      "NY/NJ, PR, VI",
      region_name
    )
  )

regions
```

## 9

Add `region` and `region_name` columns to the `population` data frame.

```{r}
population <- population |> left_join(regions, by = "state_name")

population
```

## 10

From reading <https://data.cdc.gov/> we learn the endpoint `https://data.cdc.gov/resource/pwn4-m3yp.json` provides state level data from SARS-COV2 cases. Use the **httr2** tools you have learned to download this into a data frame. Is all the data there? If not, comment on why.

```{r}
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json"

cases_raw <- httr2::request(api) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE) |>
  as_tibble()

nrow(cases_raw)
```

No. The API returns only the first 1,000 rows by default, so the result is truncated unless we explicitly increase the `$limit` parameter.

## 11

The reason you see exactly 1,000 rows is because CDC has a default limit. You can change this limit by adding `$limit=10000000000` to the request. Rewrite the previous request to ensure that you receive all the data. Then wrangle the resulting data frame to produce a data frame with columns `state`, `date` (the end date), and `cases`. Make sure the cases are numeric and the dates are in `Date` ISO-8601 format.

```{r}
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json"

cases_raw <- httr2::request(api) |>
  req_url_query(`$limit` = 10000000000) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE) |>
  as_tibble()

cases <- cases_raw |>
  transmute(
    state,
    date = parse_date_time(
      end_date,
      orders = c("ymd HMS", "ymd", "mdy HMS", "mdy"),
      quiet = TRUE
    ) |>
      as.Date(),
    cases = parse_number(new_cases)
  )

cases
```

## 12

For 2020 and 2021, make a time series plot of cases per 100,000 versus time for each state. Stratify the plot by region name. Make sure to label your graph appropriately.

```{r}
cases |>
  mutate(year = year(date)) |>
  filter(year %in% c(2020, 2021)) |>
  left_join(
    population |>
      distinct(state, year, population, region_name),
    by = c("state", "year")
  ) |>
  filter(!is.na(region_name)) |>
  mutate(cases_per_100k = cases / population * 100000) |>
  ggplot(aes(date, cases_per_100k, group = state)) +
  geom_line(alpha = 0.4, linewidth = 0.4) +
  facet_wrap(~ region_name) +
  labs(
    title = "Weekly COVID-19 cases per 100,000 by state",
    subtitle = "2020 and 2021",
    x = "Week ending date",
    y = "Cases per 100,000"
  )
```

## 13

Use the **lubridate** package to properly parse the `date` column, then create a summary table showing the total COVID-19 cases by month and year for 2020 and 2021. The table should have columns for year, month (as month name), and total cases across all states. Order by year and month. Use the **knitr** package and `kable()` function to display the results as a formatted table.

```{r}
cases |>
  mutate(date = ymd(as.character(date))) |>
  filter(year(date) %in% c(2020, 2021)) |>
  mutate(
    year = year(date),
    month_num = month(date),
    month = month(date, label = TRUE, abbr = FALSE)
  ) |>
  group_by(year, month_num, month) |>
  summarise(total_cases = sum(cases, na.rm = TRUE), .groups = "drop") |>
  arrange(year, month_num) |>
  select(year, month, total_cases) |>
  kable()
```

## 14

The following URL provides additional COVID-19 data from the CDC in JSON format:

```{r}
deaths_url <- "https://data.cdc.gov/resource/9bhg-hcku.json"
```

Use **httr2** to download COVID-19 death data from this endpoint. Make sure to remove the default limit to get all available data. Create a clean dataset called `deaths` with columns `state`, `date`, and `deaths` (renamed from the original column name). Ensure dates are in proper Date format and deaths are numeric.

```{r}
deaths_raw <- httr2::request(deaths_url) |>
  req_url_query(`$limit` = 10000000000) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE) |>
  as_tibble()

deaths <- deaths_raw |>
  filter(
    sex == "All Sexes",
    age_group == "All Ages",
    group == "By Month",
    state %in% population$state_name
  ) |>
  transmute(
    state,
    date = parse_date_time(
      end_date,
      orders = c("ymd HMS", "ymd", "mdy HMS", "mdy"),
      quiet = TRUE
    ) |>
      as.Date(),
    deaths = parse_number(covid_19_deaths)
  )

deaths
```

## 15

Using the `deaths` dataset you created, make a bar plot showing the total COVID-19 deaths by state. Show only the top 10 states with the highest death counts. Order the bars from highest to lowest and use appropriate labels and title.

```{r}
deaths |>
  group_by(state) |>
  summarise(total_deaths = sum(deaths, na.rm = TRUE), .groups = "drop") |>
  slice_max(total_deaths, n = 10) |>
  mutate(state = reorder(state, total_deaths)) |>
  ggplot(aes(state, total_deaths)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 states by total COVID-19 deaths",
    x = NULL,
    y = "Total COVID-19 deaths"
  )
```
