---
title: "Problem Set 4"
date: ""
format: html
---

[Download source](https://raw.githubusercontent.com/ywanglab/STAT3000/refs/heads/main/psets/hw4.qmd)

## Introduction

In this problem set, you will work with US Census and CDC COVID-19 data. The goal is not to produce a formal causal analysis, but rather to build a clean weekly state-level dataset and use graphics to investigate patterns in COVID-19 burden and vaccination uptake.

We will focus on the following questions:

1.  How did COVID-19 case, hospitalization, and death rates evolve during 2020 and 2021?
2.  Were there regional differences in COVID-19 burden and vaccination progress?
3.  During large national waves, did states with higher vaccination or booster coverage tend to have lower hospitalization or death rates?

## Objective

You will create a single weekly data frame for each jurisdiction in 2020 and 2021. Your final table should contain, at a minimum, the following variables:

-   `date`
-   `state`
-   `state_name`
-   `region`
-   `region_name`
-   `population`
-   `cases`
-   `hosp`
-   `deaths`
-   `vax`
-   `booster`

You will then use this table to create several figures.

## Instructions {.unnumbered}

### 1. Obtain a Census API key

Go to <https://api.census.gov/data/key_signup.html> and request a Census API key.

Do **not** hard-code your key directly into your homework file. Instead, create a file in your working directory called `_census-key.R` containing exactly one line of code:

``` r
census_key <- "YOUR_REAL_CENSUS_KEY"
```

Note that the API key must be enclosed inside the quotes. Your homework code should then load the key by sourcing that file.

### 2. Create a project structure

In your course Git repository, create a folder `hw4` that includes the following directories:

-   `data`
-   `code`
-   `figs`

Inside `code`, include:

-   `funcs.R`
-   `wrangle.R`
-   `analysis.qmd`

Your final wrangling script should save the completed dataset as an RDS file in `data`, and your figures should be saved as PNG files in `figs`.

### 3. General guidance

-   The CDC datasets do **not** all use the same variable names, so inspect the column names carefully.
-   The CDC APIs can change over time, so part of the assignment is to verify the current schema before wrangling.
-   Include comments in your code so a grader can follow your logic.
-   In each code chunk, change `#| eval: false` to `#| eval: true`.
-   Remove `NULL` in each code chunk after you add your own code.

------------------------------------------------------------------------

## Part I: Download the data

For this part, put your code in `code/wrangle.R` unless otherwise specified.

### 1. Load the required packages and source the Census key

Add code that loads the packages you will need and defines `census_key` by sourcing `_census-key.R`.

```{r}
#| eval: false
# Load packages you plan to use
#| message: false
#| warning: false

library(tidyverse)
library(httr2)
library(janitor)
library(jsonlite)
library(purrr)
library(lubridate)
library(knitr)
library(scales)

dir.create("data", showWarnings = FALSE)
dir.create("code", showWarnings = FALSE)
dir.create("figs", showWarnings = FALSE)

# Define census_key by sourcing the local file
# Your code here
source("_census-key.R")
```
## Helper functions

```{r}
#| eval: false
# Parse dates that may arrive in multiple formats
parse_date_flex <- function(x) {
  x <- as.character(x)
  out <- suppressWarnings(as_date(ymd_hms(x, quiet = TRUE))) # keep only the date
  needs_mdy <- is.na(out)
  out[needs_mdy] <- suppressWarnings(as_date(mdy(x[needs_mdy], quiet = TRUE)))
  needs_ymd <- is.na(out)
  out[needs_ymd] <- suppressWarnings(as_date(ymd(x[needs_ymd], quiet = TRUE)))
  out
}

# Pick the first matching column name from a list of candidates
pick_col <- function(dat, candidates) {
  found <- intersect(candidates, names(dat))
  if (length(found) == 0) {
    stop(
      "None of the candidate columns were found: ",
      paste(candidates, collapse = ", ") #combine them into one string and seperated by ", ". 
    )
  }
  found[1] # pick the first column
}

pick_col_or_na <- function(dat, candidates) {
  found <- intersect(candidates, names(dat))
  if (length(found) == 0) NA_character_ else found[1] # if no match, return NA_character
}

# Downloader for CDC Socrata endpoints
get_cdc_data <- function(endpoint) {
  request(endpoint) |>
    req_url_query("$limit" = 10000000) |> # limit of returned rows
    req_perform() |>  # peform  the request
    resp_check_status() |>  # 200 OK, 404: not found; 500: server error
    resp_body_json(simplifyVector = TRUE) |> #reads the response body as JSON, and simply the parse JSON into ordinary R objects when possible
    as_tibble() |>
    janitor::clean_names() #clean the column names into a standardized format (use underscore for /, -, etc)
}
```

### 2. Download Census population data

Use the Census API to request 2020 and 2021 state population estimates. The code below constructs and performs the request. Add **at least two lines of your own code** to do the following:

-   check that the response was successful,
-   determine the content type,
-   and extract the body into an object called `population_raw`.

```{r}
#| eval: false
census_url <- "https://api.census.gov/data/2021/pep/population"

population_request <- request(census_url) |>
  req_url_query(   # addes query parameters
    get = "POP_2020,POP_2021,NAME",
    "for" = "state:*",  # for=state:*: Census API syntax. state:*: every state code. "for" must be in quote for not interpreted as the reserved keyword `for`. 
    key = census_key
  )

population_response <- population_request |>
  req_perform() |>
  resp_check_status()

# Add your code below
# 1. Check the response status
# 2. Determine the content type
# 3. Extract the body into population_raw
resp_content_type(population_response)  # check the content type of the response body, here, application/json

population_raw <- population_response |>
  resp_body_json(simplifyVector = TRUE) |>
  as.data.frame(stringsAsFactors = FALSE) #keep char columns as char strings rather than factors.
```

### 3. Tidy the population data

Convert `population_raw` into a tidy tibble called `population` with one row per state-year. Your final table should include:

-   `state_name`
-   `state`
-   `year`
-   `population`

Also add state abbreviations, making sure to handle DC and Puerto Rico correctly.
```{r}
#| eval: false
population <- population_raw |>
  row_to_names(1) |> # use the first row as the column names
  as_tibble() |>
  select(NAME, POP_2020, POP_2021) |>
  rename(state_name = NAME) |>
  pivot_longer(   # wide->long
    cols = starts_with("POP_"),
    names_to = "year",
    values_to = "population"
  ) |>
  mutate(
    year = readr::parse_number(year), # convert strings like "POP_2020" into the number 2020
    population = readr::parse_number(population),
    state = case_when(
      state_name == "District of Columbia" ~ "DC",
      state_name == "Puerto Rico" ~ "PR",
      TRUE ~ state.abb[match(state_name, state.name)] # state.abb, state.name: R built-in vectors; 
    )
  ) |>
  select(state_name, state, year, population)

population
```

### 4. Download the regions file

The following JSON file lists the Public Health Service regions:

```{r}
#| eval: false
regions_url <- "https://github.com/datasciencelabs/2025/raw/refs/heads/main/data/regions.json"
```

Write code to create a data frame called `regions` with the variables:

-   `state_name`
-   `region`
-   `region_name`

One of the region names is long. Replace it with a shorter version.

```{r}
#| eval: false
regions_json <- fromJSON(regions_url, simplifyVector = FALSE)

regions <- 
  # convert the parsed list into a tidy data frame
  # shorten the longest region name
  NULL
```


```{r}
#| eval: false
regions_url <- "https://github.com/datasciencelabs/2025/raw/refs/heads/main/data/regions.json"

regions_json <- fromJSON(regions_url, simplifyVector = FALSE) # donot turn the JSON into a df or a vectors, keep it as a nested list structure. 

regions <- map_dfr(regions_json, \(x) { # used to expand nested JSON. it applies a function to each element in the list regions_json, then row-binds all results into one df
  # \(x): base R shorthand for function(x)
  tibble(
    state_name = unlist(x$states), # unlist: turns a list into a plain char vector
    region = as.integer(x$region[[1]]), #extract the first region and convert it into an integer
    region_name = x$region_name
  )
}) |>
  mutate(
    region_name = if_else(
      region_name == "New York and New Jersey, Puerto Rico, Virgin Islands",
      "NY/NJ/PR/VI",
      region_name
    )
  )

regions
```

### 5. Join population and region information

Join `regions` onto `population`.

```{r}
#| eval: false
population <- 
  # join population and regions
  NULL
```
```{r}
#| eval: false
population <- population |>
  left_join(regions, by = "state_name")

population
```

### 6. Write a reusable CDC downloader

Save the following function in `code/funcs.R`. This function should download a CDC Socrata endpoint, remove the default row limit by setting a very large `$limit`, parse the JSON, and return a tibble with clean column names.

```{r}
#| eval: false
get_cdc_data <- function(endpoint) {
  request(endpoint) |>
    req_url_query("$limit" = 10000000) |>
    req_perform() |>
    resp_check_status() |>
    resp_body_json(simplifyVector = TRUE) |>
    as_tibble() |>
    janitor::clean_names()
}
```

In `wrangle.R`, source `funcs.R`.

```{r}
#| eval: false
source("code/funcs.R")
```

### 7. Download the CDC datasets

Use the function above to download the four datasets below. Save them as `cases_raw`, `hosp_raw`, `deaths_raw`, and `vax_raw`.

-   cases: `https://data.cdc.gov/resource/pwn4-m3yp.json`
-   hospitalizations: `https://data.cdc.gov/resource/39z2-9zu6.json`
-   deaths: `https://data.cdc.gov/resource/r8kw-7aab.json`
-   vaccinations: `https://data.cdc.gov/resource/rh2h-3yt2.json`

The download code is provided. Add **one or two lines of your own** to inspect the result, for example by checking dimensions or variable names.

```{r}
#| eval: false
cases_raw  <- get_cdc_data("https://data.cdc.gov/resource/pwn4-m3yp.json")
hosp_raw   <- get_cdc_data("https://data.cdc.gov/resource/39z2-9zu6.json")
deaths_raw <- get_cdc_data("https://data.cdc.gov/resource/r8kw-7aab.json")
vax_raw    <- get_cdc_data("https://data.cdc.gov/resource/rh2h-3yt2.json")

# Add one or two quick inspection lines here
map( #apply the function dim to each elm of the list
  list(
    cases = cases_raw,
    hospitalizations = hosp_raw,
    deaths = deaths_raw,
    vaccinations = vax_raw
  ),
  dim #return number of rows, number of cols
)
```

------------------------------------------------------------------------

## Part II: Wrangle the weekly state-level dataset

In this section, you will create a weekly panel for 2020 and 2021.

### 8. Compare the raw schemas

Create a small summary table that reports, for each raw CDC dataset:

1.  the column containing the jurisdiction identifier,
2.  whether the data are daily or weekly,
3.  and the columns that contain time information.

You may render the table with `knitr::kable()`.

```{r}
#| eval: false
raw_summary <- tibble(
  outcome = c("cases", "hospitalizations", "deaths", "vaccinations")
  # add the remaining columns
)

# render the table
```
```{r}
#| eval: false
raw_summary <- tibble( # create a summary table describing the four raw datasets
  outcome = c("cases", "hospitalizations", "deaths", "vaccinations"),
  jurisdiction_column = c(
    pick_col_or_na(cases_raw, c("state", "jurisdiction", "location")),
    pick_col_or_na(hosp_raw, c("jurisdiction", "state", "location")),
    pick_col_or_na(deaths_raw, c("state", "jurisdiction")),
    pick_col_or_na(vax_raw, c("location", "state", "jurisdiction"))
  ),
  frequency = c("weekly", "daily", "weekly", "daily"), #reporting frequency of each dataset
  time_columns = c(
    paste(intersect(c("start_date", "end_date", "submission_date", "date", "date_updated"), names(cases_raw)), collapse = ", "),
    paste(intersect(c("collection_date", "date"), names(hosp_raw)), collapse = ", "),
    paste(intersect(c("start_date", "end_date", "week_ending_date", "date", "mmwr_week", "year"), names(deaths_raw)), collapse = ", "),
    paste(intersect(c("date", "mmwr_week", "mmwr_year", "date_type"), names(vax_raw)), collapse = ", ")
  )
)

kable(raw_summary)
```

### 9. Wrangle the cases data

Create a table called `cases` that contains one row per state-week with:

-   `state`
-   `mmwr_year`
-   `mmwr_week`
-   `date` (the week-ending date)
-   `cases` (weekly total cases)

Keep only states that appear in the `population` table.

Hints:

-   The date information may need to be parsed.
-   Use `epiweek()` and `epiyear()` from **lubridate**.
-   The cases dataset is already weekly, but you should still verify the time variable you use.

```{r}
#| eval: false
## Wrangle weekly cases
cases <- cases_raw |>
  # choose the relevant state column
  # choose the relevant date column
  # choose the relevant cases column
  # parse dates
  # restrict to jurisdictions in population
  # compute MMWR year and week
  # aggregate to one row per state-week if needed
  NULL
```
```{r}
#| eval: false
case_state_col <- pick_col(cases_raw, c("state", "jurisdiction", "location")) # find s which column identifies the state, and returns the first one that exists
case_date_col  <- pick_col(cases_raw, c("end_date", "submission_date", "week_ending_date", "date"))
case_n_col     <- pick_col(cases_raw, c("new_cases", "new_case", "cases"))

cases <- cases_raw |>
  transmute(  # creates a new data frame with only these three cols
    state = .data[[case_state_col]], # .data[[...]]: because the col name is stored in a variable named case_state_col; .data: current df; .data$case_state_col: Look for a col named "case_state_col". 
    date = parse_date_flex(.data[[case_date_col]]),
    cases = readr::parse_number(as.character(.data[[case_n_col]])) #extracts the numeric count
  ) |>
  semi_join(population |> distinct(state), by = "state") |> # population\>distinct(state) is the right table; distinct(state): take the state column and keep each sxtate only once.
  filter(!is.na(date)) |>
  mutate(
    mmwr_year = epiyear(date), # create epideminological year and week variables (not always the same as ordinatry canlendar year and week. THis is common for CDC-style weekly disease data)
    mmwr_week = epiweek(date)
  ) |>
  group_by(state, mmwr_year, mmwr_week) |>
  summarise(  #within each group, summarize
    date = max(date, na.rm = TRUE),
    cases = sum(cases, na.rm = TRUE),
    .groups = "drop"
  )

cases
```

### 10. Wrangle the hospitalization data

Create a table called `hosp` with the same weekly structure as `cases`, but containing weekly hospitalizations in a variable named `hosp`.

Because hospitalization data are daily, collapse them to state-week totals. Remove weeks with fewer than 7 reporting days.

```{r}
#| eval: false
## Wrangle weekly hospitalizations
hosp <- hosp_raw |>
  # identify the state, date, and hospitalization columns
  # parse the dates
  # compute MMWR year and week
  # aggregate to weekly totals
  # count number of reporting days per week
  # remove incomplete weeks
  NULL
```
```{r}
#| eval: false
hosp_state_col <- pick_col(hosp_raw, c("jurisdiction", "state", "location"))
hosp_date_col  <- pick_col(hosp_raw, c("collection_date", "date"))
hosp_n_col     <- pick_col(hosp_raw, c("new_covid_19_hospital", "new_hospital_admissions_covid"))

hosp <- hosp_raw |>
  transmute(
    state = .data[[hosp_state_col]],
    date = parse_date_flex(.data[[hosp_date_col]]),
    hosp = readr::parse_number(as.character(.data[[hosp_n_col]]))
  ) |>
  semi_join(population |> distinct(state), by = "state") |>
  filter(!is.na(date)) |>
  mutate(
    mmwr_year = epiyear(date),
    mmwr_week = epiweek(date)
  ) |>
  group_by(state, mmwr_year, mmwr_week) |>
  summarise(
    date = max(date, na.rm = TRUE),
    hosp = sum(hosp, na.rm = TRUE),
    n_days = n_distinct(date), #count distinct number of dates
    .groups = "drop"
  ) |>
  filter(n_days == 7) |>  #keep only weeks with all 7 days present
  select(-n_days) # drop the col n_days

hosp
```
### 11. Wrangle the deaths data

Create a table called `deaths` with weekly COVID-19 deaths by state. Keep the variables:

-   `state`
-   `mmwr_year`
-   `mmwr_week`
-   `date`
-   `deaths`

Be careful: this dataset may contain national rows and grouping variables.

```{r}
#| eval: false
## Wrangle weekly deaths
# Possible tasks:
# - keep only state-level observations
# - filter to a weekly grouping if needed
# - choose the COVID death count column
# - parse the week-ending date
# - keep the needed variables

deaths <- deaths_raw |>
  NULL
```

```{r}
#| eval: false
death_state_col <- pick_col(deaths_raw, c("state", "jurisdiction"))
death_date_col  <- pick_col(deaths_raw, c("week_ending_date", "end_date", "date"))
death_n_col     <- pick_col(deaths_raw, c("covid_19_deaths", "covid_deaths"))

deaths <- deaths_raw |>
  (\(x) {  # inline-anonymous function
    if ("group" %in% names(x)) filter(x, group == "By Week") else x
  })() |>  # (\(x){...}()): immediately call the function on the current dataframe
  transmute(
    state_value = .data[[death_state_col]],
    date = parse_date_flex(.data[[death_date_col]]),
    deaths = readr::parse_number(as.character(.data[[death_n_col]]))
  ) |>
  left_join(
    population |>
      distinct(state, state_name), # unique pairs of (state, state_name)
    by = c("state_value" = "state_name") # state_value in the curretn data, state_name in the populaiton table
  ) |>
  mutate(state = coalesce(state, state_value)) |> #coalesce(a,b): use a if it is not missing; otherwise use b
  select(state, date, deaths) |>
  semi_join(population |> distinct(state), by = "state") |>
  filter(!is.na(date)) |>
  mutate(
    mmwr_year = epiyear(date),
    mmwr_week = epiweek(date)
  ) |>
  group_by(state, mmwr_year, mmwr_week) |>
  summarise(
    date = max(date, na.rm = TRUE),
    deaths = sum(deaths, na.rm = TRUE),
    .groups = "drop"
  )

deaths
```
### 12. Wrangle the vaccination data

Create a table called `vax` containing the most recent vaccination record within each state-week. Keep:

-   `state`
-   `mmwr_year`
-   `mmwr_week`
-   `date`
-   `vax` (primary series cumulative count)
-   `booster` (booster cumulative count)

```{r}
#| eval: false
## Wrangle weekly vaccination data
vax <- vax_raw |>
  # identify state and date columns
  # choose cumulative primary-series and booster variables
  # parse dates
  # compute MMWR year and week
  # keep the last record within each state-week
  NULL
```
```{r}
#| eval: false
vax_state_col  <- pick_col(vax_raw, c("location", "state", "jurisdiction"))
vax_date_col   <- pick_col(vax_raw, c("date"))
vax_series_col <- pick_col(vax_raw, c("series_complete_cumulative", "series_complete_yes", "series_complete")) #completed series count
vax_boost_col  <- pick_col(vax_raw, c("booster_cumulative", "booster_doses", "additional_doses"))  # completed booster counts

vax <- vax_raw |>
  (\(x) {
    if ("date_type" %in% names(x)) filter(x, is.na(date_type) | date_type == "Report") else x
  })() |>
  transmute(
    state = .data[[vax_state_col]],
    date = parse_date_flex(.data[[vax_date_col]]),
    vax = readr::parse_number(as.character(.data[[vax_series_col]])),
    booster = readr::parse_number(as.character(.data[[vax_boost_col]]))
  ) |>
  semi_join(population |> distinct(state), by = "state") |>
  filter(!is.na(date)) |>
  mutate(
    mmwr_year = epiyear(date),
    mmwr_week = epiweek(date)
  ) |>
  arrange(state, date) |> # sort in chornological order within each state
  group_by(state, mmwr_year, mmwr_week) |>
  slice_max(date, n = 1, with_ties = FALSE) |> # for each state-week grpu, keep the row with the largest date. n: keep the n rows with the largest value. similarly, slice_head(), slice_tail()
  ungroup()

vax
```
### 13. Create the full state-week panel

We will include all state-week combinations from 2020 and 2021, even if one of the datasets is missing a value for a particular week.

Use the following scaffold to create all possible weeks and join population information:

```{r}
#| eval: false
all_dates <- data.frame(
  date = seq(make_date(2020, 1, 25), make_date(2021, 12, 31), by = "week")
) |>
  mutate(date = ceiling_date(date, unit = "week", week_start = 7) - days(1)) |>
  mutate(mmwr_year = epiyear(date), mmwr_week = epiweek(date))

dates_and_pop <- cross_join(
  all_dates,
  data.frame(state = unique(population$state))
) |>
  left_join(population, by = c("state", "mmwr_year" = "year"))
```

Now join `cases`, `hosp`, `deaths`, and `vax` to create a final table called `dat`.

Requirements:

-   order observations by `state` and `date`,
-   fill vaccination and booster counts forward within state,
-   replace missing counts for `cases`, `hosp`, and `deaths` with 0,
-   save the final table as `data/dat.rds`.

```{r}
#| eval: false
## Join the weekly tables

dat <- dates_and_pop |>
  # join weekly cases
  # join weekly hospitalizations
  # join weekly deaths
  # join weekly vaccinations
  # arrange within state by date
  # fill cumulative vaccination variables forward
  # replace missing weekly counts with 0
  NULL

## Save the dataset
# Your code here
```
```{r}
#| eval: false
all_dates <- tibble(
  date = seq(make_date(2020, 1, 25), make_date(2021, 12, 31), by = "week") # create a sequence of dates, one week apart
) |>
  mutate(
    date = ceiling_date(date, unit = "week", week_start = 7) - days(1), # adjust each date to Saturday of each week. ceiling_date(): moves each date up to the start of the next week, where weeks start on Sunday. subtract 1 day, moves back one day landing on Saturday
    mmwr_year = epiyear(date),
    mmwr_week = epiweek(date)
  )

dates_and_pop <- expand_grid(  # create all combinations of every week in all_dates adn every state in population
  all_dates,
  state = unique(population$state)
) |>
  left_join(population, by = c("state", "mmwr_year" = "year"))

dat <- dates_and_pop |>
  left_join(select(cases,  -date), by = c("state", "mmwr_year", "mmwr_week")) |>
  left_join(select(hosp,   -date), by = c("state", "mmwr_year", "mmwr_week")) |>
  left_join(select(deaths, -date), by = c("state", "mmwr_year", "mmwr_week")) |>
  left_join(select(vax,    -date), by = c("state", "mmwr_year", "mmwr_week")) |>
  group_by(state) |>
  arrange(date, .by_group = TRUE) |> #sort within each state
  fill(vax, booster, .direction = "down") |> # carry the last observed non-missing value downward to later missing rows for both vax and booster data 
  ungroup() |>
  mutate(
    cases = replace_na(cases, 0), # replacing missing values with 0
    hosp = replace_na(hosp, 0),
    deaths = replace_na(deaths, 0),
    vax = replace_na(vax, 0),
    booster = replace_na(booster, 0)
  ) |>
  arrange(state, date)

saveRDS(dat, "data/dat.rds") # save dat as an RDS file. readback : readRDS("path-to-file")

dat
```
------------------------------------------------------------------------

## Part III: Create figures in `analysis.qmd`

In your `analysis.qmd` file, load the weekly dataset from `data/dat.rds`. Each figure should have:

-   a short descriptive paragraph,
-   a code chunk that is shown in the rendered document,
-   and appropriate titles and axis labels.

```{r}
#| eval: false
# Setup chunk for analysis.qmd
library(tidyverse)
library(lubridate)
library(scales)

dat <- readRDS("data/dat.rds")
```

### 14. Figure 1: National weekly burden

Create a trend plot showing national weekly **cases**, **hospitalizations**, and **deaths** per 100,000 people. Place the three panels on top of one another.

Hint: first aggregate to the national level by week, then use `pivot_longer()` and `facet_wrap()`.

```{r}
#| eval: false
## Figure 1: National weekly burden
fig1_dat <- dat |>
  # aggregate to national weekly rates per 100,000
  NULL

# make the plot
```

### 15. Figure 2: Regional case rates over time

Create a line plot of weekly **case rates per 100,000** over time for each `region_name`. Use one line per region.

```{r}
#| eval: false
## Figure 2: Regional case rates
fig2_dat <- dat |>
  # aggregate to regional weekly rates
  NULL

# make the plot
```

### 16. Figure 3: Vaccination and booster progress by region

For each region, compute the percent of the regional population that completed the primary series and the percent that received a booster dose. Plot both over time.

One reasonable approach is to create a long data frame with one row per region-date-series.

```{r}
#| eval: false
## Figure 3: Vaccination and booster progress
fig3_dat <- dat |>
  # compute regional percentages
  NULL

# make the plot
```

### 17. Figure 4: Top 10 states by cumulative death rate

By the end of 2021, which states had the highest cumulative COVID-19 death rates per 100,000? Create a bar plot of the top 10 states.

```{r}
#| eval: false
## Figure 4: Top 10 cumulative death rates
fig4_dat <- dat |>
  # compute cumulative deaths by state
  # convert to a per-100,000 rate
  # keep the top 10
  NULL

# make the plot
```

### 18. Figure 5: Vaccination and hospitalization during a major wave

Use your earlier figures to identify **one major national COVID-19 wave in 2021** during which a meaningful share of the population had already completed the primary series.

For your chosen period:

-   compute **hospitalizations per day per 100,000** for each state,
-   compute the **primary-series vaccination rate** in each state at the end of the period,
-   create a scatter plot with vaccination rate on the x-axis and hospitalization rate on the y-axis.

Briefly explain how you chose the time window.

```{r}
#| eval: false
## Figure 5: Vaccination vs hospitalization
start_date <- as_date("YYYY-MM-DD")
end_date   <- as_date("YYYY-MM-DD")

fig5_dat <- dat |>
  # filter to the selected period
  # summarize hospitalization burden
  # join vaccination coverage at the end date
  # compute rates
  NULL

# make the plot
```

### 19. Figure 6: Booster coverage and death rates

Repeat the previous exercise, but now examine the relationship between:

-   **booster coverage** at the end of the period, and
-   **COVID-19 deaths per day per 100,000** during the period.

Use a late-2021 window for which booster coverage is nontrivial.

```{r}
#| eval: false
## Figure 6: Booster coverage vs death rate
start_date <- as_date("YYYY-MM-DD")
end_date   <- as_date("YYYY-MM-DD")

fig6_dat <- dat |>
  # filter to the selected period
  # summarize deaths during the period
  # join booster coverage at the end date
  # compute rates
  NULL

# make the plot
```

------------------------------------------------------------------------

## Deliverables

1.  Push your `hw4` folder to your Git repo that contains:

-   `code/funcs.R`
-   `code/wrangle.R`
-   `code/analysis.qmd`
-   `data/dat.rds`
-   at least six PNG files in `figs`

Submit a link to your GitHub repo. 

2.  A rendered PDF of this QMD file that should display both code and output:
```
quarto render hw4.qmd --to pdf 
```
